\vspace{3cm} \Huge{\futhark}\\ \Large{Enhancing performance by rewriting kernels from Python/OpenCL to Futhark}

### 2017-11-09 report/draft.tex
first done

\documentclass[11pt]{article}
\usepackage[a4paper, hmargin={2.8cm, 2.8cm}, vmargin={2.5cm, 2.5cm}]{geometry}
\usepackage{eso-pic} % \AddToShipoutPicture
\usepackage{graphicx} % \includegraphics
\usepackage{listings} % \includegraphics
\usepackage{pgfplots}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{mdframed}
\usepackage{float}
\usepackage{amsmath}
\floatstyle{boxed}
\restylefloat{figure}

@@

\newcommand{\sasmodels}{\textit{SasModels}}
\newcommand{\sasview}{\textit{SasView}}
\newcommand{\iq}{\textit{I(q)}}
\newcommand{\foriq}{\texttt{for\_iq}}
\newcommand{\hasiq}{\texttt{has\_iq}}
\newcommand{\futhark}{\texttt{Futhark for SasModels}}
\newcommand{\barheight}{5cm}
\newcommand{\call}{\texttt{\_\_call\_\_}}

@@

\clearpage\maketitle
\thispagestyle{empty}
\clearpage
\section*{Abstract}
\sasmodels{} performs small angle scattering calculations for the \sasview{}
modelling suite, by sending it's input to a range of calculation models
implemented in Python and OpenCL. 
\\
A Futhark module was written and added to the 
existing \sasmodels{} codebase, to test whether Futhark could outperform the 
already existing model languages used in \sasmodels{}. The Futhark models
were performance tested against a range of the original models.
All following results showed a speed increase of between 200 and 2000\%, 
all without loss of precision.  Speedups were however only apparent at 
unrealistically large input sizes.
\\
Future \sasmodels{} extensions are then suggested, so these speed increases can be 
leveraged in daily \sasview{} usage.

@@

\section{Introduction}
\sasview{} is a Small Angle Scattering (SAS) analysis package for the
analysis of 1D and 2D scattering data directly in inverse space.
When \sasview{} is given input data, it performs its analyses by running the
input data through theory models provided by the standalone module
\sasmodels{}.
\\\\
The \sasmodels{} models utilizes a set of kernels, which are implemented
in both Python and OpenCL. However, benchmark
comparisons\cite{futhark}
between the parallel algorithm library \textit{Thrust}
(implemented in C++/CUDA), and the array programming language Futhark, shows
that there might severe speed improvements to gain by running these model
calculations in Futhark, instead of in OpenCL as they are now.
\\\\
The goal of this project is to explore the potential speed gains obtained by
rewriting, and testing the resulting speed improvements of, a subset of the
\sasmodels{} models.
The subset will be chosen so that there are for each of Python and OpenCL 
one ``simple'' model, and a complex one.
\\\\
If rewriting the models results in a significant\footnote{I.e. a speed increase
  in order of at least 100\%}, speedup, it will reasonable to believe that 
  the rest of the models in \sasmodels{} can be rewritten for a similar
  speedup.

\section{Demonstration}
\begin{mdframed}[
  frametitle={Terminology:},
  nobreak=true
  ]
  The work of this project has been spread out across multiple parts
  of the \sasmodels{} program. Here is a list of the distinct parts of 
  \sasmodels{} that has been worked with:
  \begin{itemize}
    \item \texttt{kernelfut}: The \sasmodels{} Futhark kernel module that wraps 
    \sasmodels{} input data and sends it to be calculated in the actual Futhark 
    kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
    \texttt{kernelcl} which are already available in \sasmodels.

@@

    \item Futhark models: A Futhark model is the model module that contains the
    functions needed to run a given model. See the Futhark line model in fig. 
    \ref{fig:linemodel-futhark}

@@

All \sasmodels{} models are described in model info file. These model info 
files has at least two properties:
\begin{itemize}
  \item a function definition \texttt{I(q)}, which defines scattered intensity 
  of the scattering vector \texttt{q}.

  \item a list \texttt{parameters}, of the parameters that are used in 
  \texttt{I(q)}. If \texttt{I(q)} does not take any parameters besides 
  \texttt{q}, the list is left empty.
  In the model info file, the parameters are each defined with their unit name, 
  their default values, their value range, their type of parameter and
  a short text describing what they are.
\end{itemize}

Additionally, the model info file also always contains the short-name, 
the full title, a category, and a mathematical \footnote{a representation 
that compiles to a meaningful mathematical expression if used in a LaTeX math 
environment} string representation of the scattering function.
\\\\
As the scattering function \texttt{I(q)} and its parameters are the only 
two things that are strictly necessary in a model definition, the developer
has the option to leave the model info defined with just these six properties.
\\\\
\sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
for working with voluminous scatterings.
\\\\
Therefore, it is optional for the developer to define a custom definition
of $I(q_x, q_y)$ or other default functions.
An example of this can be seen in fig. \ref{fig:linemodel}.
A more complex model can be seen in fig. \ref{fig:core_shell}

\subsection{Defining \iq{} in Python and OpenCL}
A \sasmodels{} \iq{} function can be written in either Python (which will be
executed in a Numpy context at runtime), or it can be written in C, which will
will be inserted into an OpenCL template, so that it can be compiled and 
executed as a kernel in an OpenCL context, preferably on a GPU.
\\\\
A Python \iq{} can be seen in figure \ref{fig:linemodel}, whereas both DAB 
(figure
\ref{fig:dabmodel} and core\_shell(figure \ref{fig:core_shell}) are written in C.
Notice that the main difference between the models lies in the Iq property.
For Python models, \iq{} is a callable Python function, whilst the C \iq{} is a
 string which is copy-pasted wholesale into aforementioned OpenCL template.
\\\\
But with the addition of Futhark models in \sasmodels{} adds a third way of
using and defining a \sasmodels{} model.

\subsection{Using a Futhark model}
\label{sec:using-futhark}
There is a difference between defining and using a Futhark model.
The \sasview{} user who wishes to use a Futhark model in his or her work does 
not need more than the futhark model info file, and a copy of the compiled
Futhark kernel. In fact, the futhark model info file is identical to OpenCL- 
and Python model info files, except for in two properties.
\\\\
As shown in figure \ref{fig:linemodelinfo-futhark}, \iq{} is now a dictionary
with the single flag \texttt{model\_is\_futhark} set to \texttt{True},
and the property \texttt{futhark\_path} points to the location of the compiled
Futhark kernel.
\\\\
The user does not need to have any Futhark compiler installed if he merely wants
to use the Futhark kernel: \futhark{} compiles to portable Python libraries which 
has numpy and pyopencl as their only dependencies - and as \sasmodels{} has these 
dependencies as well, the user will already have these dependencies sorted out, 
if he has a working installation of \sasview{} or \sasmodels{}.

\subsection{Defining a Futhark model}
\label{sec:defining-futhark-model}
To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
A model \textit{m} itself is defined as a parametric Futhark module\footnote{
  A more in-depth explanation of the complete definition of a Futhark module is 
  in \ref{sec:generating-futhark}},
which contains an Iq-function, an Iqxy- and a form\_volume function.
It is the model programmers responsibility to fill in all three functions.
\\\\
A simple example of a Futhark model is the line module, 
as depicted in figure \ref{fig:linemodel-futhark}.
Note that Iqxy is non-default, whilst form\_volume is the default 
implementation.

\subsubsection{On the user friendliness of Futhark models}
As shown in figure \ref{fig:linemodel-futhark}, there are currently two small 
nuisances in Futhark models, compared to equivalent models written in
Python/OpenCL:
\\
With both Python and OpenCL models, the programmer can leave out Iqxy and 
form\_volume so that \sasmodels{} automatically fills in the default definition
for these two functions instead.
\\\\
The Futhark programmer does not have this leisure yet, and must define the two
functions no matter what. This is admittedly a piece of annoying boiler plate,
but it is also a thing that can be solved by compartmentalising the offending
functions away in a parametric library module.
\\\\
The less removable nuisance is that the programmer must explicitly extract the
Iq function parameters from the list \texttt{local\_values}. This must be done
in the order they appear in the model info parameter list.
Iq functions in Python and OpenCL can be defined with any number of function 
parameters, but in Futhark, we need to adhere to the type declaration of Iq 
(see \texttt{has\_iq} definition in \ref{fig:for-iq}), which means that Iq can 
take only two arguments; the scattering vector \texttt{q}, and the list 
\texttt{local\_values} of parameters needed for the scattering.
\\\\
This makes for an ugly prelude to the actual body of the Iq function, as shown
in fig\ref{fig:linemodel-futhark}, or more critically in models like 
core-shell \ref{fig:core_shell}.

\subsection{Generating Futhark models using parametric modules and a header}
\label{sec:generating-futhark}
As mentioned in \ref{sec:defining-futhark-model}, the futhark model developer
does only need to define three functions to be able to compile a working 
\sasmodels{} Futhark model.
However, those three functions are merely used by a parametric Futhark module 
\foriq{}, which takes care of the actual execution of the 
Futhark kernel.
Let's go through the actual implementation of the Futhark models.

@@

There a several things happening in this fig. \ref{fig:linemodel-futhark-full}, 
besides the definition of the line Iq function.
\\
\texttt{module line} takes either single- or double precision floats as a
parameter, and then assigns the type alias \texttt{dtype} to the type of the
input parameter. This means that we can use \texttt{dtype} in the type
declarations throughout the module, and define the module for floats in general,
instead of maybe having to define linef64- and linef32 modules.
\\\\
Then, two kernels (one for each of f32 and f64) are initialised by giving
the parametric module \texttt{for\_iq} the line module and a data type 
as parameters, returning \texttt{kernel\_float32} and \texttt{kernel\_float64}.
\\
Given these two modules, we can finally define four entry points into the
finished model. These are the four entry points that can be called by 
\sasmodels{} whenever a \sasmodels{} experiment needs to execute calculations 
in a Futhark kernel.
Each of the entry points are simply passing on their call parameters to the 
\texttt{run\_kernel} functions defined in the \texttt{for\_iq} module.
Everything after the module line definition is boilerplate code, and could
possibly defined in a less inelegant way in a later revision, but currently, it 
is a necessary part of every Futhark model.
\\\\
Other models, such as core-shell imports libraries so they are 
available in the Iq function.

\subsubsection{header/for\_iq.fut (the Futhark kernel template)}
\texttt{for\_iq.fut} contains the module type declaration for the \hasiq{} module,
 and it contains the definition of the parametric \foriq{} module. 
\\
The type declaration \hasiq{} ensures, that any model module that is
used as a parameter for the \foriq{} module defines at least the functions
Iq, Iqxy and form\_volume, as these functions are necessary for the definition
of the Futhark kernel.
\\\\
Next, we have the actual definition of the module \foriq{}:
\\
\foriq{} takes two parameters (a parametric Futhark model), and a real type 
to instantiate it with.
It then instantiates the model with the real type, making the finished Iq module
\texttt{Iq} available throughout the rest of the kernel.
\\\\
Then finally, it uses the Iq module to define the two functions 
\texttt{run\_kernel}, and \texttt{run\_kernel\_2d}, which defines the actual 
\sasmodels{} computation procedure with the scattering functions injected by 
the Iq module.

\begin{figure}
  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
  \caption{The Futhark template that is used by the individual models.}
  \label{fig:for-iq}
\end{figure}

\subsubsection{Why not implement Futhark models as model info strings 
instead of standalone modules?}
OpenCL kernels are generated by copy/pasting the desired \iq function body
into a template, and then compiling the resulting source code.
This happens at runtime in \sasmodels, which means that the kernels aren't
 actually compiled until right before \sasmodels{} needs them.
\\\\
Futhark models eschews this design and opts instead to separate code generation
and model usage completely.
It would be simple to copy the design of the OpenCL kernels, and compile Futhark
kernels on \sasmodels{} runtime, but this design was cast aside in favour of the
design described in sec \ref{sec:generating-futhark}.
\\
There are three reasons for that:
\begin{enumerate}
  \item Removing code generation from runtime unequivocally decreases the time
  it takes to run a \sasmodels{} calculation. Although time taken by code 
  generation is not covered by \sasmodels' built-in benchmark modules, it should
  be obvious that a program that merely executes a kernel computation has a 
  shorter runtime than a program that has to compile the kernel before it
  executes the kernel computation.

@@

To design \texttt{kernelfut}, it was first necessary to ape the 
\sasmodels{} calculation loop as a Futhark function. 
I chose to replicate \texttt{kernelpy}'s \texttt{\_loops(...)} function in 
Futhark, as it is fairly simple to rewrite the Python code in Futhark.
\texttt{\_loops()} is shown in figure \ref{fig:kernelpy_loop}, and the
comparable Futhark function in figure \ref{fig:kernelfut_loop}.
\\\\
Second, it was necessary to implement a bridge between \sasmodels{} and the 
precompiled Futhark kernels. 
When \sasmodels{} wants to use a Futhark kernel for it's calculations, 
it must dynamically import the kernel during the initialisation of the FutKernel
object. Figure \ref{fig:retrieving-futhark-class} shows how FutKernel uses
the path and model name from the model info file, to load and initialize the
model class that is available in the compiled Futhark kernel.
\\\\
After the Futhark kernel has been initialised and made available to FutKernel,
we can retrieve the kernel's OpenCL queue, and buffer our input data onto the 
OpenCL device immediately, so we don't have to spend time doing this during 
kernel execution itself.
Doing this reduces the Futhark overhead during kernel execution by between
200 and 800 milliseconds.
\\\\
Finally, the FutKernel's _form attribute is set. _form is the entry point from
the Futhark kernel, as seen in figure \ref{fig:linemodel-futhark-full}.

%% what is form??
\begin{figure}
  \lstinputlisting[linerange={30-32, 55-56, 74-82, 121-126}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
  \label{fig:retrieving-futhark-class}
  \caption{Injecting the precompiled Futhark kernel in the FutKernel object}
\end{figure}

\begin{figure}
  \lstinputlisting[linerange={128-140}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
  \label{fig:buffering-kernelfut}
  \caption{Buffering the input data to the Futhark kernel's queue}
\end{figure}

Like with \texttt{kernelcl} and \texttt{kernelpy}, FutKernel must also implement
\call{} so that DirectModel can call the model and perform the
calculations.
This has been done by writing a simplified version of the OpenCL kernel's 
\call{} function. 
\\\\
As seen in figure \ref{fig:kernelfut_call} and \ref{fig:kernelcl_call},
the primary task of the call function is to pack an array of arguments, and
pass them on as arguments for a kernel.
As \call{} must return a numpy array, the get() function is called at the end
of the Futhark kernel call, to convert the output OpenCL Array into a 
numpy array.
\\\\
Note that kernelfut does not have the evaluation loop that kernelcl has.
The loop still exists, but has been moved inside of the Futhark template 
instead (see fig. \ref{fig:for-iq}).

\section{Method}
\sasmodels{} contains a comparison module called \texttt{compare.py}, which
lets the user run different calculations on the same dataset.
This module has been the basis of all my performance tests, and indeed also
the module that has let me design Futhark models in the first place.
\\\\
The development of Futhark for \sasmodels{} has been largely test-driven.
For any given already existing model in the \sasmodels{} catalogue that I have
wished to implement in Futhark, I have started out by defining an equivalent
model written in Futhark, and then used the comparison module to compare the
 runs of the original model versus the Futhark implementation.
\\\\
The comparison module displays the average execution time of each of the models,
which lets me track exactly how much faster (or slower) the Futhark model has
been running, compared to the original implementation.
\\\\
Furthermore, and ultimately most important, the comparison module also
revealed the difference in results between the two models. 
As I have precisely zero knowledge on the field of Small Angle Scatterings, 
the only option for me regarding correctness of the models has been to take
the results of the already existing models as the "correct" results of the
computations. Consequently, any difference between my Futhark results and the
original results has been direct indicators of faulty implementations or bugs
somewhere along my work.
Therefore, I have been developing and debugging my solution by repeatedly
comparing it's results to that of the original implementation.
\\\\
Conversely, I have chosen to accept test runs with exactly no difference
between the original and the Futhark implementation, as indicators of a
correct implementation. 
Test runs with thousands of data points defined as double-precision
with more than ten decimal points have still resulted in a total difference of
zero\footnote{see \ref{sec:correctness}} between the models.
\\
This is what I base my assertion of a correctness on, as even the smallest 
mistakes along the calculations would surely have resulted in a very significant
accumulated error in the final results.
\\\\
The test driven development has also lead to some of the biggest performance
optimizations in my design of kernelfut.py.
In example, the initial tests showed that the Futhark kernels were unreasonably
much slower than the already existing kernels. After some investigation I found
that this was because the original implementation buffered the input data in the
OpenCL context long time before the actual kernel was called in Python, whereas
the first versions of the Futhark kernel buffered the input data in the OpenCL
context as part of the \texttt{run\_kernel} function call.

@@

      \end{axis}
    \end{tikzpicture}
    \label{fig:line-graph}
    \caption{Comparison between Python and Futhark performance for simple model}
\end{figure}

In figure \ref{fig:line-graph} we see that for one dimensional calculations,
both implementations keep a steady execution time. Futhark is consistently 
slower, but that can be explained by the overhead of using the Futhark kernels.
\\\\
More interesting is it to look at the execution time of the 2D version of the 
experiment. For all \sasmodels{} 2D experiments, we are not just calculating \iq
for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
So if we have $n$ q, we need to perform $n^2$ calculations.
\\\\
We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
the Python model.

>>>>

\documentclass[11pt]{article}
\usepackage[a4paper, hmargin={2.8cm, 2.8cm}, vmargin={2.5cm, 2.5cm}]{geometry}
\usepackage{eso-pic} % \AddToShipoutPicture
\usepackage{graphicx} % \includegraphics
\usepackage{listings} % \includegraphics
\usepackage{pgfplots}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{mdframed}
\usepackage{float}
\usepackage{amsmath}
\usepackage{flafter}
\floatstyle{boxed}
\restylefloat{figure}

@@

\newcommand{\sasmodels}{\textit{SasModels}}
\newcommand{\sasview}{\textit{SasView}}
\newcommand{\iq}{\texttt{I(q)}}
\newcommand{\foriq}{\texttt{for\_iq}}
\newcommand{\hasiq}{\texttt{has\_iq}}
\newcommand{\futhark}{\texttt{Futhark for SasModels}}
\newcommand{\barheight}{5cm}
\newcommand{\call}{\texttt{\_\_call\_\_}}

@@

\clearpage\maketitle
\thispagestyle{empty}
\clearpage
\section*{Abstract}
\sasmodels{} performs small angle scattering calculations for the \sasview{}
modeling suite, by sending its input to a range of calculation models
implemented in Python and OpenCL. 
\\
A Futhark module was written and added to the 
existing \sasmodels{} codebase, to test whether Futhark could outperform the 
already existing model languages used in \sasmodels{}. The Futhark models
were performance tested against a range of the original models.
All following results showed a speed increase of between 200 and 2000\%, 
all without loss of precision.  Speedups were however only apparent at 
unrealistically large input sizes.
\\
Future \sasmodels{} extensions are then suggested, so these speed increases can be 
leveraged in daily \sasview{} usage.

@@

\section{Introduction}
\sasview{} is a Small Angle Scattering (SAS) analysis package for the
analysis of 1D and 2D scattering data directly in inverse space.
When \sasview{} is given input data, it performs its analyses by running the
input data through theoretical models provided by the standalone module
\sasmodels{}.
\\\\
\sasmodels{} utilizes a set of kernels, which are implemented
in both Python and OpenCL. However, benchmark
comparisons\cite[sec. 5.3]{array16}
between the parallel algorithm library \textit{Thrust}
(implemented in C++/CUDA), and the array programming language Futhark, shows
that there might be severe speed improvements to gain by running these model
calculations in Futhark, instead of in OpenCL as they are now.\\
\\\\
The goal of \futhark{} is to explore the potential speed gains obtained by
rewriting, and testing the resulting speed improvements of, a subset of the
models in \sasmodels{}.
The subset will be chosen so that there are for each of Python and OpenCL 
one ``simple'' model, and a complex one.
\\\\
If rewriting the models results in a significant\footnote{I.e. a speed increase
  in order of at least 100\%} speedup, it will be reasonable to believe that 
  the rest of the models in \sasmodels{} can be rewritten to obtain a similar
  speedup.

\section{Demonstration of the implementation}
\begin{mdframed}[
  frametitle={Terminology:},
  nobreak=true
  ]
  \futhark{} is spread across multiple parts
  of the \sasmodels{} program. Here is a list of the distinct parts of 
  \sasmodels{} that has been worked with:
  \begin{itemize}
    \item \texttt{kernelfut}: The \sasmodels{} Futhark kernel module that wraps 
    \sasmodels{} input data and sends it to be calculated in the actual Futhark 
    kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
    \texttt{kernelcl} which are already available in \sasmodels.

@@

    \item Futhark models: A Futhark model is the model module that contains the
    functions needed to run a given model. See the Futhark line model in figure 
    \ref{fig:linemodel-futhark}

@@

All \sasmodels{} models are described in model info file. These model info 
files has at least two properties:
\begin{itemize}
  \item a function definition \iq{}, which defines scattered intensity 
  of the scattering vector \texttt{q}.

  \item a list \texttt{parameters}, of the parameters that are used in 
  \iq{}. If \iq{} does not take any parameters besides 
  \texttt{q}, the list is left empty.
  In the model info file, the parameters are each defined with their unit name, 
  their default values, their value range, their type of parameter and
  a short text describing what they are.
\end{itemize}

Additionally, the model info file also always contains the short-name, 
the full title, a category, and a mathematical \footnote{a representation 
that compiles to a meaningful mathematical expression if used in a LaTeX math 
environment} string representation of the scattering function.
\\\\
As the scattering function \iq{} and its parameters are the only 
two things that are strictly necessary in a model definition, the developer
has the option to leave the model info defined with just these six properties.
\\\\
\sasmodels{} has built-in default definitions for \texttt{I(qx,qy)} if working
with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
for working with voluminous scatterings.
\\\\
Therefore, it is optional for the developer to define a custom definition
of \texttt{I(qx, qy)} or other default functions.
An example of this can be seen in figure \ref{fig:linemodel}.

\subsection{Defining \iq{} in Python and OpenCL}
A \sasmodels{} \iq{} function can be written in either Python (which will be
executed in a Numpy context at runtime), or it can be written in C, which will
will be inserted into an OpenCL template, so that it can be compiled and 
executed as a kernel in an OpenCL context, preferably on a GPU.
\\\\
A Python \iq{} can be seen in figure \ref{fig:linemodel}, whereas \iq{} for
both DAB 
(figure \ref{fig:dabmodel}) and core\_shell(figure \ref{fig:core_shell})
are written in C.
Notice that the main difference between the models lies in the Iq property.
For Python models, \iq{} is a callable Python function, whilst the C \iq{} is a
 string which is copy-pasted wholesale into the aforementioned OpenCL template.
\\
With \futhark{}, a third way is introduced for using and defining a \sasmodels{} model.

\subsection{Using a Futhark model}
\label{sec:using-futhark}
There is a difference between defining and using a Futhark model.
The \sasview{} user who wishes to use a Futhark model in his or her work does 
not need more than the model info file, and the Futhark kernel.
A Futhark model info file is identical to OpenCL- 
and Python model info files, except for in two properties.
\\\\
As shown in figure \ref{fig:linemodelinfo-futhark}, \iq{} is now a dictionary
with the single flag \texttt{model\_is\_futhark} set to \texttt{True},
and the property \texttt{futhark\_path} points to the location of the Futhark kernel.
\\\\
The user does not need to have the Futhark compiler installed if he merely wants
to use the Futhark kernel: \futhark{} compiles to portable Python libraries which 
has Numpy and Pyopencl as their only dependencies - and as \sasmodels{} has these 
dependencies as well, the user will already have these dependencies sorted out, 
if he has a working installation of \sasview{} or \sasmodels{}.

\subsection{Defining a Futhark model}
\label{sec:defining-futhark-model}
To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
A Futhark model \textit{m} itself is defined as a parametric Futhark module\footnote{
  A more in-depth explanation of the complete definition of a Futhark module is 
  in sec \ref{sec:generating-futhark}},
which contains an Iq-function, an Iqxy- and a form\_volume function.
It is the model programmers responsibility to fill in all three functions.
\\\\
A simple example is the line module, shown in figure \ref{fig:linemodel-futhark}.
Note that Iqxy is a non-default definition, whilst form\_volume is the default 
definition.

\subsubsection{On the user friendliness of Futhark models}
As apparent from figure \ref{fig:linemodel-futhark}, there are currently two small 
nuisances in Futhark models, compared to equivalent models written in
Python/OpenCL:
\\
With both Python and OpenCL models, the programmer can leave out Iqxy and 
form\_volume so that \sasmodels{} automatically fills in the default definition
for these two functions instead.
\\\\
The Futhark programmer does not have this leisure, but must therefore always
define the two functions. This is boiler plate code, but in the future it could
be solved by compartmentalising the offending functions away in a
parametric library module.
\\\\
Less fixable is it that the programmer must explicitly extract the
Iq function parameters from the list \texttt{local\_values}. This must be done
in the order they appear in the model info parameter list.
\\
Iq functions in Python and OpenCL can be defined with any number of function 
parameters, but Futhark models must adhere to the type declaration of Iq 
(see \texttt{has\_iq} definition in figure \ref{fig:for-iq}), which means that Iq can 
take only two arguments; the scattering vector \texttt{q}, and the list 
\texttt{local\_values} of parameters needed for the scattering.
\\\\
This leads to long preludes to the actual function body Iq function, as in
figure \ref{fig:linemodel-futhark} (line model), or more glaringly like in
figure \ref{fig:core_shell} (core shell model).

\subsection{Generating Futhark models using parametric modules and a header}
\label{sec:generating-futhark}
As mentioned in sec. \ref{sec:defining-futhark-model}, the Futhark model developer
does only need to define three functions to be able to compile a working 
\sasmodels{} Futhark model.
However, those three functions are merely used by a parametric Futhark module 
\foriq{}, which takes care of the actual execution of the 
Futhark kernel.
Let's go through the actual implementation of the Futhark models.

@@

Several things are happening in line\_futhark.fut besides (figure
\ref{fig:linemodel-futhark-full}), besides the definition of the line Iq function.
\\
\\
\texttt{module line} takes either single- or double precision floats as a
parameter, and then assigns the type alias \texttt{dtype} to the type of the
input parameter. This means that we can use \texttt{dtype} in the type
declarations throughout the module, and define the module for floats in general,
instead of maybe having to define linef64- and linef32 modules.
\\\\
Then, two kernels (one for each of f32 and f64) are initialised by giving
the parametric module \texttt{for\_iq} the line module and a data type 
as parameters, returning \texttt{kernel\_float32} and \texttt{kernel\_float64}.
\\
Given these two modules, we can finally define four entry points into the
finished model. These are the four entry points that can be called by 
\sasmodels{} whenever a \sasmodels{} experiment needs to execute calculations 
in a Futhark kernel.
Each of the entry points are simply passing on their call parameters to the 
\texttt{run\_kernel} functions defined in the \texttt{for\_iq} module.
Everything after the module line definition is boilerplate code, and could
possibly be defined in a more elegant way in a later revision, but currently, it 
is a necessary part of every Futhark model.
\\\\
Other models, such as core shell, imports libraries which are then
available in the Iq function.

\subsubsection{header/for\_iq.fut (the Futhark kernel template)}
The file \texttt{for\_iq.fut} contains the module type declaration for the \hasiq{} module,
 and it contains the definition of the parametric \foriq{} module. 
\\
The type declaration \hasiq{} ensures, that any model module that is
used as a parameter for the \foriq{} module defines at least the functions
Iq, Iqxy and form\_volume, as these functions are necessary for the definition
of the Futhark kernel.
\\\\
Next, we have the actual definition of the module \foriq{}, which takes two parameters (a parametric Futhark model), and a real type to instantiate it with.
It then instantiates the model with the real type, making the finished Iq module
\texttt{Iq} available throughout the rest of the kernel.
\\
Then finally, it uses the Iq module to define the two functions 
\texttt{run\_kernel}, and \texttt{run\_kernel\_2d}, which defines the actual 
\sasmodels{} computation procedure with the scattering functions injected by 
the Iq module.

\begin{figure}
  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
  \caption{The Futhark template that is used by the individual models. Notice
    that this template is not copied, but appears only once in the system - it
    is shared between the Futhark models.}
  \label{fig:for-iq}
\end{figure}

\subsubsection{Why not implement Futhark models as model info strings 
instead of standalone modules?}
OpenCL kernels are generated by copy/pasting the desired \iq function body
into a template, and then compiling the resulting source code.
This happens at runtime in \sasmodels, which means that the kernels aren't
 actually compiled until right before \sasmodels{} needs them.
\\\\
Futhark models eschews this design and opts instead to separate code generation
and model usage completely.
It would be simple to copy the design of the OpenCL kernels, and compile Futhark
kernels at \sasmodels{} runtime, but this design was cast aside in favour of the
design described in sec \ref{sec:generating-futhark}.
\\
There are three reasons for this decision:
\begin{enumerate}
  \item Removing code generation from runtime unequivocally decreases the time
  it takes to run a \sasmodels{} calculation. Although time taken by code 
  generation is not covered by \sasmodels' built-in benchmark modules, it should
  be obvious that a program that merely executes a kernel computation has a 
  shorter runtime than a program that has to compile the kernel before it
  executes the kernel computation.

@@

To design \texttt{kernelfut}, it was first necessary to ape the 
\sasmodels{} calculation loop as a Futhark function. 
I chose to replicate \texttt{kernelpy}'s \texttt{\_loops(...)} function in 
Futhark, as it is fairly simple to rewrite the Python code in Futhark.
\texttt{\_loops()} is shown in figure \ref{fig:kernelpy_loop}, and the
comparable Futhark function in figure \ref{fig:kernelfut_loop}.
\\\\
Second, it was necessary to implement a bridge between \sasmodels{} and the 
precompiled Futhark kernels. 
When \sasmodels{} wants to use a Futhark kernel for it's calculations, 
it must dynamically import the kernel during the initialisation of the FutKernel
object. Figure \ref{fig:retrieving-futhark-class} shows how FutKernel uses
the path and model name from the model info file, to load and initialize the
model class that is available in the compiled Futhark kernel.
\\\\
After the Futhark kernel has been initialised and made available to FutKernel,
we can retrieve the kernel's OpenCL queue, and buffer our input data onto the 
OpenCL device immediately, so we don't have to spend time doing this during 
kernel execution itself.
Doing this reduces the Futhark overhead during kernel execution by between
200 and 800 milliseconds.
\\\\
Finally, the FutKernel's \texttt{\_form} attribute is set. \texttt{\_form} is the entry point from
the Futhark kernel, as seen in figure \ref{fig:linemodel-futhark-full}.

%% what is form??
\begin{figure}
  \lstinputlisting[linerange={30-32, 55-56, 74-82, 121-126}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
  \caption{Injecting the precompiled Futhark kernel in the FutKernel object}
  \label{fig:retrieving-futhark-class}
\end{figure}

\begin{figure}
  \lstinputlisting[linerange={128-140}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
  \caption{Buffering the input data to the Futhark kernel's queue}
  \label{fig:buffering-kernelfut}
\end{figure}

Like with \texttt{kernelcl} and \texttt{kernelpy}, FutKernel must also implement
\call{} so that DirectModel can call the model and perform the
calculations.
This has been done by writing a simplified version of the OpenCL kernel's 
\call{} function. 
\\\\
As seen in figure \ref{fig:kernelfut_call} and figure \ref{fig:kernelcl_call},
the primary task of the call function is to pack an array of arguments, and
pass them on as arguments for a kernel.
As \call{} must return a numpy array, the get() function is called at the end
of the Futhark kernel call, to convert the output OpenCL Array into a 
numpy array.
\\\\
Note that kernelfut does not have the evaluation loop that kernelcl has.
The loop still exists, but has been moved inside of the Futhark template 
instead (see figure \ref{fig:for-iq}).

\section{Method}
\sasmodels{} contains a comparison module called \texttt{compare.py}, which
lets the user run different calculations on the same dataset.
This module has been the basis of all my performance tests, and indeed also
the module that has let me design Futhark models in the first place.
\\\\
The development of Futhark for \sasmodels{} has been largely test-driven.
For any given already existing model in the \sasmodels{} catalogue that I have
wished to implement in Futhark, I have started out by defining an equivalent
model written in Futhark, and then used the comparison module to compare the
 runs of the original model versus the Futhark implementation.
\\\\
The comparison module displays the average execution time of each of the models,
which lets me track exactly how much faster (or slower) the Futhark model has
been running, compared to the original implementation.
\\\\
Furthermore, and ultimately most important, the comparison module also
revealed the difference in results between the two models. 
As I have limited knowledge about the field of Small Angle Scatterings, 
the only option for me regarding correctness of the models has been to take
the results of the already existing models as the "correct" results of the
computations. Consequently, any difference between my Futhark results and the
original results has been direct indicators of faulty implementations or bugs
somewhere in my work.
Therefore, I have been developing and debugging my solution by repeatedly
comparing it's results to that of the original implementation.
\\\\
Conversely, I have chosen to accept test runs with exactly no difference
between the original and the Futhark implementation, as indicators of a
correct implementation. 
Test runs with thousands of data points defined as double-precision
with more than ten decimal points have still resulted in a total difference of
zero\footnote{see sec. \ref{sec:correctness}} between the models.
\\
This is what I base my assertion of a correctness on, as even the smallest 
mistakes along the calculations would surely have resulted in a very significant
accumulated error in the final results.
\\\\
The test driven development has also lead to some of the biggest performance
optimizations in my design of kernelfut.py.
In the example, the initial tests showed that the Futhark kernels were unreasonably
slower than the already existing kernels. After some investigation I found
that this was because the original implementation buffered the input data in the
OpenCL context long time before the actual kernel was called in Python, whereas
the first versions of the Futhark kernel buffered the input data in the OpenCL
context as part of the \texttt{run\_kernel} function call.

@@

      \end{axis}
    \end{tikzpicture}
    \caption{Comparison between Python and Futhark performance for simple model}
    \label{fig:line-graph}
\end{figure}

In figure \ref{fig:line-graph} we see that for one dimensional calculations,
both implementations keep a steady execution time. Futhark is consistently 
slower, but that can be explained by the overhead of launching the Futhark kernels.
\\\\
More interesting is it to look at the execution time of the 2D version of the 
experiment. For all \sasmodels{} 2D experiments, we are not just calculating \iq
for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
So if we have a q vector of $n$ elements, we need to perform $n^2$ calculations.
\\\\
We see in the figure, that Futhark at $10000^2$ datapoints now greatly outperforms
the Python model.

\subsubsection{\texttt{broad\_peak} model}
\texttt{broad\_peak} is quite different than the line model from before (see
figure \ref{fig:broadpeakmodel}). Its \iq
function is defined as 
$I(q) = \frac{A}{q^n} + \frac{C}{1 + (|q - q_0|\xi)^m} + B$.\footnote{see \url{http://www.sasview.org/docs/user/models/broad_peak.html}}
